VarSight: Prioritizing Clinically Reported Variants with Binary Classification Algorithms

### 2019-01-24 paper/main.tex
added a preprint toggle for biorxiv and easy switch to normal

\documentclass{bioinfo}
\usepackage{caption}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{hhline}
\copyrightyear{2015} \pubyear{2015}

@@

@@

\begin{methods}
\section{Methods}
\subsection{Data sources}
All samples were selected from the cohort of Undiagnosed Diseases Network (UDN) \citep{ramoni2017} genome sequencing samples that were originally sequenced at HudsonAlpha Institute for Biotechnology (HAIB).  In short, the UDN accepts patients with rare, undiagnosed diseases that are believed to have a genetic origin.  The UDN is not restricted to a particular disease, so there are a diverse set of diseases and phenotypes represented across the whole population.  The phenotypes annotated to a patient are also noisy compared to simulated datasets for a variety of reasons including: 1) some patients have multiple diseases, 2) phenotype collection is done at seven different clinical sites leading to slightly different standards of collection, and 3) some patients exhibit more or fewer phenotypes than are associated with the classic disease presentation.  For more details on the UDN, refer to \citealp{ramoni2017}.

@@

We then selected four readily available models for classification that are capable of training on imbalanced datasets: a random forest model by {\it sklearn} \citep{pedregosa2011}, a logistic regression model by {\it sklearn}, a balanced random forest model by {\it imblearn} \citep{lemaitre2017}, and an ensemble classification model by {\it imblearn}.  For each model, we selected a list of hyperparameters to test and tested each possible combination of those hyperparameters.  For each model and set of hyperparameters, we performed 10-fold cross validation on the training variants and recorded the balanced accuracy scores (i.e. a weighted accuracy score where inputs are weighted by their inverse class frequency).  For each model type, we saved the hyperparameters and model with the best average balanced accuracy score.  These four tuned models were then tested against the unseen set of test proband cases.

\end{methods}

@@

For each tuned model, we calculated the 10-fold cross validated balanced accuracy on the training set.  We then calculated the true positive rate (TPR), false positive rate (FPR), and area under the receiver operator curve (AUROC) based on the unseen test data (see Table \ref{tab:classifier_results}.  
%The values for each model are shown in Table \ref{tab:classifier_results}.  
Figure \ref{fig:roc} show receiver operator curves for FPR v. TPR and recall v. precision for the four models.  %Figure \ref{fig:roc} also shows the receiver operator curves for two single-measure statistics (CADD Scaled and HPO-cosine) that are used in the next sections as comparisons for ranking variants.

@@

\begin{table}[!t]
\processtable{Classifier performance statistics.  For each tuned classifier, we show performance measures commonly used for classifiers (from left to right): 10-fold cross validation balanced accuracy (CV10 Acc.), true positive rate (TPR), false positive rate (FPR), and area under the receiver operator curve (AUROC). In general, the two random forest methods and the EasyEnsembleClassifier perform very similarly whereas LogisticRegression seems to be slightly worse across all measures. \label{tab:classifier_results}}{
\begin{tabular}{c|c|c|c|c}
\hline
\input{classifier_rendered.tex}
\hline
\end{tabular}}{}
\end{table}

\subsection{Ranking Statistics}
\begin{table*}[!t]
\centering
\processtable{Ranking performance statistics.  This table shows the ranking performance statistics for all methods evaluated on our test set.  CADD Scaled and HPO-cosine are single value measures that were used as inputs to the classifiers we tested.  Exomiser is an external tool that only reported ranks for a subset of the filtered variants.  ``Exomiser(best)" conservatively assumed unranked primary variants were ranked at the next best position despite being unranked.  In contrast, ``Exomiser(avg)" realistically assumed unranked primary variants were at the average rank (i.e. middle) for all unranked variants.  The bottom four rows are the tuned, binary classification methods tested in this paper.  The ``Case Rank" columns show the median and mean ranks for all reported variants along with the variants split into their reported pathogenicity (calculated using the ACMG guidelines).  The ``Percentage in Top X Variants" columns show the percentage of variants that were found in the top 1, 10, and 20 variants in a case after ranking by the corresponding method.  All values were generated using only the test data that was unseen during training. \label{tab:performance}}{
\begin{tabular}{c|c|c|c|c||c|c|c|c}
\hline
\input{data_rendered.tex}
\hline
\end{tabular}}{}%{$^*$Unranked variants were conservatively assumed to be at the next best rank after those that were ranked.}
\end{table*}

@@

\begin{table}[!t]
\processtable{Random forest feature importances.  This table shows the top feature importances reported by the two random forest algorithms we tested.  We had exactly 50 features passed into the classifier algorithms, so we show all features with an average importance $\ge 0.02$ indicating a feature utilized more than we would expect by chance. \label{tab:features}}{
\begin{tabular}{l|c|c}
\hline
\input{features_rendered.tex}
\hline
\end{tabular}}{}
\end{table}

>>>>

\newif\ifpreprint
\preprinttrue

\ifpreprint
\documentclass{article}
\else
\documentclass{bioinfo}
\fi

\usepackage{caption}
\usepackage{graphicx}
\usepackage{hhline}
\usepackage{multirow}

\ifpreprint
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{amsmath}
%need these to rotate
\usepackage{geometry}
\usepackage{pdflscape}
\usepackage{scrextend}
\usepackage{authblk}

\begin{document}
\title{VarSight: Prioritizing Clinically Reported Variants with Binary Classification Algorithms}
\author[1]{James M. Holt}
\author[1]{Brandon Wilk}
\author[1]{Camille L. Birch}
\author[1]{Donna M. Brown}
\author[1]{Manavalan Gajapathy}
\author[1]{Alexander C. Moss}
\author[1]{Nadiya Sosonkina}
\author[1]{Melissa A. Wilk}
\author[1]{Julie A. Anderson}
\author[1]{Jeremy M. Harris}
\author[1]{Jacob M. Kelly}
\author[1]{Fariba Shaterferdosian}
\author[1]{Angelina E. Uno-Antonison}
\author[1]{Arthur Weborg}
\author[ ]{Undiagnosed Diseases Network}
\author[1]{Elizabeth A. Worthey}
\affil[1]{Software Development and Informatics, HudsonAlpha Institute for Biotechnology, Huntsville, 35806, USA}
\maketitle

\begin{abstract}
\noindent\textbf{Motivation:} In genomic medicine for rare disease patients, the primary goal is to identify one or more variants that cause their disease.  Typically, this is done through filtering and then prioritization of variants for manual curation.  However, prioritization of variants in rare disease patients remains a challenging task due to the high degree of variability in phenotype presentation and molecular source of disease.  Thus, methods that can identify and/or prioritize variants to be clinically reported in the presence of such variability are of critical importance.  \\
\textbf{Results:} We tested the application of classification algorithms that ingest variant predictions along with phenotype information for predicting whether a variant will ultimately be clinically reported and returned to a patient.  To test the classifiers, we performed a retrospective study on variants that were clinically reported to 237 patients in the Undiagnosed Diseases Network.  We treated the classifiers as variant prioritization systems and compared them to another variant prioritization algorithm and two single-measure controls. We showed that these classifiers outperformed the other methods with the best classifier ranking 73\% of all reported variants and 97\% of reported pathogenic variants in the top 20.
\\
\textbf{Availability:} The scripts used to generate results presented in this paper are available at \href{https://github.com/HudsonAlpha/VarSight}{https://github.com/HudsonAlpha/VarSight}. \\
\textbf{Contact:} \href{jholt@hudsonalpha.org}{jholt@hudsonalpha.org}\\
\end{abstract}

\else%begin non-preprint

\copyrightyear{2015} \pubyear{2015}

@@

%end non-preprint
\fi

@@

\ifpreprint

\else
\begin{methods}
\fi
\section{Methods}
\subsection{Data sources}
All samples were selected from the cohort of Undiagnosed Diseases Network (UDN) \citep{ramoni2017} genome sequencing samples that were originally sequenced at HudsonAlpha Institute for Biotechnology (HAIB).  In short, the UDN accepts patients with rare, undiagnosed diseases that are believed to have a genetic origin.  The UDN is not restricted to a particular disease, so there are a diverse set of diseases and phenotypes represented across the whole population.  The phenotypes annotated to a patient are also noisy compared to simulated datasets for a variety of reasons including: 1) some patients have multiple diseases, 2) phenotype collection is done at seven different clinical sites leading to slightly different standards of collection, and 3) some patients exhibit more or fewer phenotypes than are associated with the classic disease presentation.  For more details on the UDN, refer to \citealp{ramoni2017}.

@@

We then selected four readily available models for classification that are capable of training on imbalanced datasets: a random forest model by {\it sklearn} \citep{pedregosa2011}, a logistic regression model by {\it sklearn}, a balanced random forest model by {\it imblearn} \citep{lemaitre2017}, and an ensemble classification model by {\it imblearn}.  For each model, we selected a list of hyperparameters to test and tested each possible combination of those hyperparameters.  For each model and set of hyperparameters, we performed 10-fold cross validation on the training variants and recorded the balanced accuracy scores (i.e. a weighted accuracy score where inputs are weighted by their inverse class frequency).  For each model type, we saved the hyperparameters and model with the best average balanced accuracy score.  These four tuned models were then tested against the unseen set of test proband cases.
\ifpreprint

\else
\end{methods}
\fi

@@

For each tuned model, we calculated the 10-fold cross validated balanced accuracy on the training set.  We then calculated the true positive rate (TPR), false positive rate (FPR), and area under the receiver operator curve (AUROC) based on the unseen test data (see Table \ref{tab:classifier_results}).  
%The values for each model are shown in Table \ref{tab:classifier_results}.  
Figure \ref{fig:roc} show receiver operator curves for FPR v. TPR and recall v. precision for the four models.  %Figure \ref{fig:roc} also shows the receiver operator curves for two single-measure statistics (CADD Scaled and HPO-cosine) that are used in the next sections as comparisons for ranking variants.

@@

\ifpreprint
\begin{table}[!t]
\centering
\begin{tabular}{c|c|c|c|c}
\hline
\input{classifier_rendered.tex}
\hline
\end{tabular}
\caption{Classifier performance statistics.  For each tuned classifier, we show performance measures commonly used for classifiers (from left to right): 10-fold cross validation balanced accuracy (CV10 Acc.), true positive rate (TPR), false positive rate (FPR), and area under the receiver operator curve (AUROC). In general, the two random forest methods and the EasyEnsembleClassifier perform very similarly whereas LogisticRegression seems to be slightly worse across all measures.}
\label{tab:classifier_results}
\end{table}

\else

\begin{table}[!t]
\processtable{Classifier performance statistics.  For each tuned classifier, we show performance measures commonly used for classifiers (from left to right): 10-fold cross validation balanced accuracy (CV10 Acc.), true positive rate (TPR), false positive rate (FPR), and area under the receiver operator curve (AUROC). In general, the two random forest methods and the EasyEnsembleClassifier perform very similarly whereas LogisticRegression seems to be slightly worse across all measures. \label{tab:classifier_results}}{
\begin{tabular}{c|c|c|c|c}
\hline
\input{classifier_rendered.tex}
\hline
\end{tabular}}{}
\end{table}
\fi

\ifpreprint
\newgeometry{margin=1cm} % modify this if you need even more space
\begin{landscape}
\begin{table}[!t]
\centering
\begin{tabular}{c|c|c|c|c||c|c|c|c}
\hline
\input{data_rendered.tex}
\hline
\end{tabular}
\caption{Ranking performance statistics.  This table shows the ranking performance statistics for all methods evaluated on our test set.  CADD Scaled and HPO-cosine are single value measures that were used as inputs to the classifiers we tested.  Exomiser is an external tool that only reported ranks for a subset of the filtered variants.  ``Exomiser(best)" conservatively assumed unranked primary variants were ranked at the next best position despite being unranked.  In contrast, ``Exomiser(avg)" realistically assumed unranked primary variants were at the average rank (i.e. middle) for all unranked variants.  The bottom four rows are the tuned, binary classification methods tested in this paper.  The ``Case Rank" columns show the median and mean ranks for all reported variants along with the variants split into their reported pathogenicity (calculated using the ACMG guidelines).  The ``Percentage in Top X Variants" columns show the percentage of variants that were found in the top 1, 10, and 20 variants in a case after ranking by the corresponding method.  All values were generated using only the test data that was unseen during training.}
\label{tab:performance}
\end{table}
\end{landscape}
\restoregeometry
\subsection{Ranking Statistics}

\else

\subsection{Ranking Statistics}
\begin{table*}[!t]
\centering
\processtable{Ranking performance statistics.  This table shows the ranking performance statistics for all methods evaluated on our test set.  CADD Scaled and HPO-cosine are single value measures that were used as inputs to the classifiers we tested.  Exomiser is an external tool that only reported ranks for a subset of the filtered variants.  ``Exomiser(best)" conservatively assumed unranked primary variants were ranked at the next best position despite being unranked.  In contrast, ``Exomiser(avg)" realistically assumed unranked primary variants were at the average rank (i.e. middle) for all unranked variants.  The bottom four rows are the tuned, binary classification methods tested in this paper.  The ``Case Rank" columns show the median and mean ranks for all reported variants along with the variants split into their reported pathogenicity (calculated using the ACMG guidelines).  The ``Percentage in Top X Variants" columns show the percentage of variants that were found in the top 1, 10, and 20 variants in a case after ranking by the corresponding method.  All values were generated using only the test data that was unseen during training. \label{tab:performance}}{
\begin{tabular}{c|c|c|c|c||c|c|c|c}
\hline
\input{data_rendered.tex}
\hline
\end{tabular}}{}%{$^*$Unranked variants were conservatively assumed to be at the next best rank after those that were ranked.}
\end{table*}
\fi

@@

\ifpreprint
\begin{table}[!t]
\centering
\begin{tabular}{l|c|c}
\hline
\input{features_rendered.tex}
\hline
\end{tabular}
\caption{Random forest feature importances.  This table shows the top feature importances reported by the two random forest algorithms we tested.  We had exactly 50 features passed into the classifier algorithms, so we show all features with an average importance $\ge 0.02$ indicating a feature utilized more than we would expect by chance.}
\label{tab:features}
\end{table}

\else

\begin{table}[!t]
\processtable{Random forest feature importances.  This table shows the top feature importances reported by the two random forest algorithms we tested.  We had exactly 50 features passed into the classifier algorithms, so we show all features with an average importance $\ge 0.02$ indicating a feature utilized more than we would expect by chance. \label{tab:features}}{
\begin{tabular}{l|c|c}
\hline
\input{features_rendered.tex}
\hline
\end{tabular}}{}
\end{table}
\fi
